### 2015-09-01 deprecated/pattern-match-analysis/thesis/relatedwork.tex
pattern-match-analysis

\documentclass[a4paper]{report}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{alltt}

\usepackage{tikz}

\begin{document}

\title{Thesis Proposal}
\author{Ruud Koot}
\maketitle

% Also see:
% http://www.cs.uu.nl/wiki/bin/view/Thesis/WritingAThesisProposal
% mail from Jur
% Tamar's/Eric's proposal

\chapter*{Introduction}

This document is a proposal for a thesis on answering the research question:
\begin{quote}
Can we use a type and effect system in combination with refinement types to develop a pattern-match analysis for a non-strict higher-order functional language that is both performant and precise enough to be of practical use?
\end{quote}

In Chapter \ref{secmotivation} we present a number of examples to demonstrate why this is an interesting problem. In Chapter \ref{seccontext} we give a short introduction to the relevant concepts of the research question: higher-order functional languages, type and effect systems and refinement types. In Chapter \ref{secrelatedwork} we present work related to our research question and discuss which aspects of that work are relevant to or different from our proposed system. In Chapter \ref{secapproach}, \ref{secdeliverables} and \ref{secevaluation} we will discuss our envisioned approach to answering the research question, the set of deliverables and how we will evaluate whether the research question has been answered successfully or not.

\chapter{Motivation}\label{secmotivation}

In 1978, Robin Milner \cite{Milner78atheory} famously wrote (and proved) that:
\begin{quotation}
Well-typed programs cannot ``go wrong''. 
\end{quotation}
This observation is sometimes
%\footnote{Especially by professors wishing to impress their audience of undergraduate students with the power of types.}
optimistically overstated as ``if your Haskell program type checks and compiles it will work.'' Even if we ignore logic errors, such a strong interpretation of Milner's statement does not hold, as anyone who has seen the dreaded:
\begin{verbatim}
*** Exception: PM.hs:1:1-13: Non-exhaustive patterns in function f
\end{verbatim}
can attest to. To correctly interpret Milner we need to distinguish between three kinds of ``wrongness'':
\begin{description}
\item[Getting stuck] If a program tries to evaluate a non-sensical expression -- such as {\tt 3 + true} -- it cannot possibly make any further progress and is said to be ``stuck''. This is the ``go wrong'' Milner referred to. He proved that a sound type system can statically guarantee such expressions will never occur or need to be evaluated at run-time.
\item[Diverging] If we have a function definition {\tt f = f}, the evaluation of {\tt f} $\leadsto$ {\tt f} $\leadsto ...$ will fail to terminate. We might be making progress in a technical sense, but it will have no useful observable result.
\item[Undefinedness] Another source of ``wrongness'' are partial functions: functions which are not defined on all the elements of their domain. Prime examples are case-statements with missing constructors and functions defined by pattern matching but which do not cover all possible patterns. 
\end{description}

Unlike Milner, who spoke about the first kind of wrongness and the work on termination checking, which concerns itself with the second, we shall focus on the third: a pattern-match analysis which determines if functions are only invoked on values for which they are defined.

% split error and partial definedness?
% loop and crash

\section{Examples}\label{secexamples} % test-driven development

\paragraph{Partial functions} Haskell programmers often work with partial functions, the most common one possibly being {\tt head}:

\begin{verbatim}
1  main = let xs = if length "foo" > 5 then [1,2,3] else []
2          in head xs
3
4  head (x : xs) = x
\end{verbatim}

This program is guaranteed to crash at run-time, so we would like to be warned beforehand by the compiler or a tool:

\begin{verbatim}
On line 2 you applied the function "head" to the empty list "xs".
The function "head" expects a non-empty list as its first argument.
\end{verbatim}

If the guard of the if-statement had read {\tt length "foo" < 5} the program would have run without crashing and we would like the compiler or tool not to warn us spuriously. In case it is not possible to determine statically whether or not a program will crash, a warning should still be raised.

\paragraph{Compiler construction}
Compilers work with large and complex data types to represent the abstract syntax tree. These data structures must be able to represent all syntactic constructs the parser is able to recognize. This results in an abstract syntax tree that is unnecessarily complex and too cumbersome for the later stages of the compiler -- such as the optimizer -- to work with. This problem is resolved by \emph{desugaring} the original abstract syntax tree into a simpler -- but semantically equivalent --  abstract syntax tree than does not use all of the constructors available in the original abstract syntax tree.

The compiler writer now has a choice between two different options: either write a desugaring stage {\tt desugar :: ComplexAST -> SimpleAST} -- duplicating most of the data type representing and functions operating on the abstract syntax tree -- or take the easy route {\tt desugar :: AST -> AST} and assume certain constructors will no longer be present in the abstract syntax tree at stages of the compiler executed after the desugaring step. The former has all the usual downsides of code duplication -- such as having to manually keep the two or more data types synchronized -- while that latter forgoes many of the advantages of strong typing and type safety: if the compiler pipeline is restructured and one of the stages that was originally assumed to run only after the desugaring suddenly runs before that point the error might only be detected at run-time by a pattern match failure. A pattern match analysis should be able to detect such errors statically.

\paragraph{Maintaining invariants}
Many algorithms and data structures maintain invariants that cannot be easily encoded into their type. These invariants often ensure that certain incomplete case-statements are guaranteed not to cause a pattern match failure. An example is the {\tt risers} function from \cite{DBLP:conf/sfp/MitchellR05}, calculating monotonically increasing segments of a list (e.g., {\tt risers [1,3,5,1,2]} $\leadsto$ {\tt [[1,3,5],[1,2]]}):

\begin{verbatim}
1  risers :: Ord a => [a] -> [[a]]
2  risers []         = []
3  risers [x]        = [[x]]
4  risers (x1:x2:xs) = let (s:ss) = risers (x2:xs)
5                       in if x <= y then (x:s):ss else [x]:(s:ss)
\end{verbatim}

The let-binding in the third alternative of {\tt risers} expects the recursive call to return a non-empty list. A naive analysis might raise a warning here. If we think a bit longer, however, we see that we also pass the recursive call to {\tt risers} a non-empty list. This means we will end up in either the second or third alternative in the recursive call. Both the second alternative and both branches of the if-statement in the third alternative result in a non-empty list, satisfying the assumption we made earlier.

Another example might be a collection of mathematical operations working on bitstrings (integers encoded as lists of binary digits without leading zeroes):
\begin{verbatim}
1  type Bitstring = [Int]
2
3  add :: Bitstring -> Bitstring -> Bitstring
4  add [] y = y
5  add x [] = x
6  add (0:x) (0:y) = 0 : add x y
7  add (0:x) (1:y) = 1 : add x y
8  add (1:x) (0:y) = 1 : add x y
9  add (1:x) (1:y) = 0 : add (add [1] x) y
\end{verbatim}

The patterns in {\tt add} are far from complete, but maintain the invariant if passed arguments that satisfy the invariant. So if we are careful to only pass valid bitstrings into a complex mathematical expression of bitstring-operations it will result in a valid bitstring without crashing due to a pattern match failure.

% Examples from Refinement Types for ML:
% * Binary encoded numbers
% * Variable substitutions in Boolean expressions

\chapter{Context}\label{seccontext}

% halting problem
% well-typed programs can't go wrong
% smaller types log form presentation

\section{Haskell}

Haskell is statically-typed functional programming language with non-strict evaluation semantics \cite{Haskell98}. As a functional language it has first-class and higher-order functions and features a rich type system supporting parametric polymorphism and type classes. Programmers can define custom types in the form of algebraic data types and write functions over them using pattern matching.

Like most functional languages, Haskell can be easily translated into a typed $\lambda$-calculus (System $F_C$). From the point-of-view of the programmer it offers a wealth of syntactic sugar over a plain $\lambda$-calculus -- such as guards and list comprehensions -- allowing programs to be expressed concisely and in a readable fashion.

The following example demonstrates the the syntax of Haskell and some features mentioned previously:

\begin{verbatim}
1  data Tree a = Branch (Tree a) (Tree a)
2              | Leaf a
3  
4  mapTree :: (a -> b) -> Tree a -> Tree b
5  mapTree f (Branch t1 t2) = Branch (mapTree f t1) (mapTree f t2)
6  mapTree f (Leaf a)       = Leaf (f a)
7
8  instance Functor Tree where
9      fmap = mapTree
\end{verbatim}
Lines 1--2 define an algebraic data type (ADT) representing a tree. A tree can be constructed by either a {\tt Branch} containing a left and a right subtree or by a {\tt Leaf} containing a value. The ADT is parameterized by a type {\tt a} giving the type of the value stored in the {\tt Leaf}.

Lines 4--5 define a higher-order function {\tt mapTree} which applies a given function {\tt f :: a -> b} to all the values stored in all the leaves of a given {\tt Tree} resulting in a new {\tt Tree}. As the domain and the range of {\tt f} need not coincided the type parameter of the tree changes as well. The body of the function is defined by pattern matching over the constructors of {\tt Tree} and performs recursion in the case of a {\tt Branch} and applies the argument {\tt f} to the field of type {\tt a} in {\tt Leaf} otherwise.

Lines 8--9 declare the type constructor {\tt Tree} to be an instance of the type class {\tt Functor}, expressing it to be an endofunctor in the category of data types with its mapping on morphisms given by {\tt mapTree}.

\subsection{Pattern matching}

Particularly relevant to our analysis are Haskell's pattern matching abilities. We demonstrate some of them in the following function:

\begin{verbatim}
rotate (Branch b@(Branch (Leaf x) (Leaf y)) (Leaf z))
           | z < x     = Branch (Leaf z) b
           | z < y     = Branch (Branch (Leaf x) (Leaf z)) (Leaf y)
           | otherwise = undefined
\end{verbatim}

The function will rotate a tree {\tt (Branch (Branch (Leaf 2) (Leaf 3)) (Leaf 1))} into {\tt Branch (Leaf 1) (Branch (Leaf 2) (Leaf 3))}. It accomplishes this through the use of \emph{nested pattern matching}, \emph{as-patterns} and \emph{guards}.

This function only works on trees of a very specific shape. Feeding the function a tree of a different shape will cause a pattern match failure. Similarly, an error will be raised if the function reaches the {\tt otherwise} branch, containing an {\tt undefined} value. This would also happen if the branch had called {\tt error} or the branch and its guard were left out entirely.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Type and effect systems}

Type and effect systems are approaches to program analysis suitable for typed languages \cite{Nielson:1999:PPA:555142}. In this formalism we take the the \emph{underlying type system} of the language and (conservatively) extend it by adding \emph{annotations} to the types (base, function and other) of the system.

\begin{figure}[p]
\[\frac{}{\Gamma \vdash_{\mathrm{UL}} c : \tau_c}\ \mathrm{[con]} \quad\quad\quad\quad
\frac{\Gamma(x) = \tau}{\Gamma \vdash_{\mathrm{UL}} x : \tau}\ \mathrm{[var]}\]
\[\frac{\Gamma[x \mapsto \tau_1] \vdash_{\mathrm{UL}} e : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \lambda x. e : \tau_1 \to \tau_2}\ \mathrm{[abs]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_2 \to \tau_1 \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} e_1\ e_2 : \tau_1}\ \mathrm{[app]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_1 \quad \Gamma[x\mapsto \tau_1] \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \tau_2}\ \mathrm{[let]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau}{\Gamma \vdash \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau}\ \mathrm{[if]}\]
\caption{Underlying type system}
\label{underlying}
\end{figure}

\begin{figure}[p]
\[\frac{}{\Gamma \vdash_{\mathrm{AN}} n : \mathrm{int}^{\{n\}}}\ \mathrm{[con]} \quad\quad\quad\quad
\frac{\Gamma(x) = \widehat\tau}{\Gamma \vdash_{\mathrm{AN}} x : \widehat\tau}\ \mathrm{[var]}\]
\[\frac{\Gamma[x \mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \lambda x. e : \widehat\tau_1 \to \widehat\tau_2}\ \mathrm{[abs]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_2 \to \widehat\tau_1 \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} e_1\ e_2 : \widehat\tau_1}\ \mathrm{[app]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_1 \quad \Gamma[x\mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \widehat\tau_2}\ \mathrm{[let]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_2} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_3}}{\Gamma \vdash_{\mathrm{AN}} \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau^{\varphi_2 \cup \varphi_3}}\ \mathrm{[if]}\]
\[\frac{
\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi}
}{
\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi \cup \varphi'}
}\ \mathrm{[sub]}\]
\caption{Annotated type system}
\label{annotated}
\end{figure}

\subsection{Example}

As an example we give an analysis which determines, in the form of an annotation, which values an expression can evaluate to. The language is the simply-typed $\lambda$-calculus with let-bindings and an if-statement. The underlying type systems is standard and given in Figure \ref{underlying}.

The annotated type system is given in Figure \label{annotated}. It does little more than collecting all constants in the annotations. The only interesting rule is for the if-statement. We need to take the union over the values collected in both branches of the if-statement. We could have chosen to make the rule depend on the annotation of the guard expression, passing on only the values collected in the then-branch along if we knew it could only be {\tt True} or only the values collected in the else-branch if we knew it could only be {\tt False} from its annotation. This would make the analysis more precise.

We have also added an inference rule for \emph{subeffecting}. This allows us to make the annotation less precise. This is sometimes necessary when applying a value about which we have very accurate information to a function which requires a more general argument.

\subsection{Type inference}

An analysis specified as an inference system does not allow us to compute the result of the analysis directly, as the premises of several of the inference rules usually require us to guess the correct type annotations for the type and annotation variables (e.g., in the presence of subeffecting). To overcome this problem we need a \emph{type reconstruction algorithm}.

If the underlying type system is Hindley--Milner, then Algorithm W can be used to infer the types. Algorithm W works bottom-up and can compute types in a single pass by unifying type variables. This may not always be possible for the annotations in an annotated type system, however. Here we can use a variant of Algorithm W to gather a set of constraints on the annotation variables and solve these constraints with a worklist algorithm in a second phase.

\subsection{Polyvariance}

% 1) precision (poisoning)
% 2) modularity

% annotation polymorphism

Consider the following program:
\begin{verbatim}
f = let id x = x
     in (id 1, id 2)
\end{verbatim}

When performing the analysis given above we would assign the type $\mathtt{int} \to \mathtt{int}^{\{1, 2\}}$ to {\tt id} and as a result the type $\mathtt{int}^{\{1,2\}} \times \mathtt{int}^{\{1,2\}}$ to {\tt f}. Clearly, this result is not optimal. This loss in precision has been caused by the two separate calls to {\tt id} \emph{poisoning} each other.

A similar situation occurs at the type-level in the function:
\begin{verbatim}
f = let id x = x
     in (id 1, id true)
\end{verbatim}
Hindley--Milner manages to avoid problems in this situation by a mechanism called \emph{let-generalization}. The binding {\tt id} has the type $\alpha \to \alpha$. Instead of trying -- and failing -- to unify $\alpha$ with both {\tt int} and {\tt bool} in the body of the let-binding, the type of {\tt id} is generalized to the polymorphic type $\forall \alpha. \alpha \to \alpha$. In the body of the let-binding, {\tt id} can be instantiated twice: once to $\mathtt{int} \to \mathtt{int}$ and once to $\mathtt{bool} \to \mathtt{bool}$.

A similar trick can be used for the annotated type system, except that instead of quantifying over type variables, we quantify over annotation variables.

\section{Refinement types}

\begin{figure}[t]
\centering
\begin{tikzpicture}
    \tikzstyle{all nodes}=[inner sep=4pt]
    \draw node(top)at(1,3){\tt List a}
          node(sorn)at(1,2){{\tt Singleton a} $\lor$ {\tt Nil a}}
          node(s)at(0,1){\tt Singleton a} node(n)at(2,1){\tt Nil a}
          node(bot)at(1,0){$\bot$};
    \draw[-](top)--(sorn);
    \draw[-](sorn)--(s);
    \draw[-](sorn)--(n);
    \draw[-](s)--(bot);
    \draw[-](n)--(bot);
\end{tikzpicture}
\caption{A lattice of {\tt List} and its {\tt rectype}s.}
\label{listlattice}
\end{figure}

Refinement types are a form of subtyping that allow us to state more accurately what values a variable of a particular type can contain. For example, a refinement type {\tt \{ a :: Int | a < 0 \}} states that {\tt a} is a variable of type {\tt Int} than can only contain negative numbers.

@@

Their key contribution is allowing the programmer to define {\tt rectype}s. Besides defining a recursive type for lists:

\begin{verbatim}
data List a = Nil | Cons a (List a)
\end{verbatim}

The programmer can also define a {\tt rectype} describing singleton lists:

@@

Together with an automatically derived {\tt rectype} for the non-recurisive constructor {\tt Nil} (and a union type contructor) the compiler can construct -- using known algorithms on regular tree grammars \cite{DBLP:books/others/tree1984} -- a finite lattice of refinement types for {\tt List a} given in Figure \ref{listlattice}.

\subsection{Union and Intersection Types}

@@

A union type such as {\tt Int $\lor$ Bool}, {\tt List a $\lor$ Singleton a} or {\tt Int -> Int $\lor$ Bool -> Bool} state a value can be either of the type on the left or of the type on the right, but we do not know which. In the case of {\tt List a $\lor$ Singleton a} we are able to simplify this type to {\tt List a}, as {\tt Singleton a} is a subtype of {\tt List a}.

@@

\begin{verbatim}
Cons :: a -> Nil       a -> Singleton a
     /\ a -> Singleton a -> List      a
     /\ a -> List      a -> List      a
\end{verbatim}

This type can be derived automatically from the {\tt rectype} declarations. Types of functions can also be inferred automatically, although there may be a loss of precision when higher-order functions or polymorphism are involved.

@@

where {\tt case\_list} has the refinement type

@@

\chapter{Related Work}\label{secrelatedwork}

Catch works by calculates preconditions on functions. The preconditions are given in a constraint language. By varying the constraint language trade-offs between performance and precision can be made. If the calculated precondition for \texttt{main} is True, the program does not crash.

Preconditions are calculated iteratively. Precondition start out as \texttt{True}, except for \texttt{error}, whose precondition is \texttt{False}.\footnote{Note that {\tt undefined} is defined in terms of {\tt error}.}

\paragraph{Example} Following the example given by Mitchell, given the function:
\begin{verbatim}
safeTail xs = case null xs of
                True  -> []
                False -> tail xs
\end{verbatim}
The computed precondition will be:
\[ \textrm{Precondition}(\texttt{null xs}) \land (\texttt{null xs} \in \{\texttt{True}\} \lor \textrm{Precondition}(\texttt{tail xs})), \]
stating the necessary precondition for \texttt{null xs} not to crash mush be fulfilled and either \texttt{null xs} must evaluate to \texttt{True} or the necessary precondition for \texttt{tail xs} not to crash must be fulfilled.

As \texttt{null xs} cannot crash we find its precondition to be simply True and we can deduce the precondition for \texttt{tail xs} to be $\texttt{xs}\in\{(:)\}$. Substituting these into our precondition we find
\[ \textrm{True} \land (\texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}), \]
which simplifies to
\[ \texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}. \]
The subexpression  $\texttt{null xs} \in \{\texttt{True}\}$ forms a postcondition on \texttt{null xs}. From this postcondition Catch computes the required precondition that satisfies it, which in this particular case should turn out to be $\texttt{xs}\in\{[]\}$. Substituting, we find
\[ \texttt{xs}\in\{[]\} \lor \texttt{xs}\in\{(:)\}, \]
which, using our knowledge of the list data type, turns out the be a tautology, i.e. the (trivial) precondition for \texttt{safeTail} is True.

\subsection{Constraint systems}
The example in the previous section used a simple constraint language, specifying to which set of head constructors an expression should evaluate. Mitchell developed two more expressive constraint systems: regular expression constraints and multipattern constraints.

\paragraph{Regular expression constraints} Constraints are formed by a regular expression over an alphabet of a number of ad-hoc \emph{selectors} (e.g. \texttt{hd} and \texttt{tl} for lists.) A precondition for \texttt{map head xs} using regular expression constraints reads $\texttt{xs} \in (\texttt{tl}^{*} \cdot \texttt{hd} \leadsto \{(:)\})$. It should be interpreted as ``after applying zero or more \texttt{tl}s to \texttt{xs} and then applying a \texttt{hd} we should find a \texttt{(:)} constructor.''

Accoriding to Mitchell regular expression constraints tend to scale badly as the program increase in size, although he could not identify more specific condition under which this problem manifests.\footnote{Personal communication}

\paragraph{Multipattern constraints} Multipatterns are of the form $\alpha \star \rho$, where $\alpha$ gives the set of constructors that are valid at the head of the value, while $\rho$ gives the set of constructors that can appear at the recursive positions (if any) of the constructor at the head of the value. The elements of these sets can again be multipatterns. To specify that \texttt{xs} in \texttt{map head xs} should be a list of non-empty lists, we use the multipattern:
\begin{align*}
\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\} \\
&\star \\
\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\}
\end{align*}
\texttt{Any} is a wildcard that matches any constructor.

Programmers define pre- and postconditions in the form of a contract (a refinement type). An appropriate contract for the \texttt{head} function defined above might be:
\begin{verbatim}
{-# CONTRACT head :: {s | not (null s)) -> {x | True} #-}
\end{verbatim}
This contract states that if head is given a list \texttt{s} for which the predicate \texttt{not (null s)} holds, i.e. it is a non-empty list, the function will not crash\footnote{The system only guarantees partial correctness, so the term might still diverge.}, indicated by the trivial postcondition \texttt{True} on the return value.

>>>>

\chapter{Related Work}\label{chaprelatedwork}

Catch works by calculates preconditions on functions. The preconditions are given in a constraint language. By varying the constraint language trade-offs between performance and precision can be made. If the calculated precondition for |main| is True, the program does not crash.

Preconditions are calculated iteratively. Precondition start out as |True|, except for |error|, whose precondition is |False|.\footnote{Note that |undefined| is defined in terms of |error|.}

\paragraph{Example} Following the example given by Mitchell, given the function:
\begin{code}
safeTail xs =  case null xs of
                  True   -> []
                  False  -> tail xs
\end{code}
the computed precondition will be:
\[ \textrm{Precondition}(|null xs|) \land (|null xs| \in \{|True|\} \lor \textrm{Precondition}(|tail xs|)), \]
stating the necessary precondition for |null xs| not to crash must be fulfilled and either |null xs| must evaluate to |True| or the necessary precondition for |tail xs| not to crash must be fulfilled.

As |null xs| cannot crash we find its precondition to be simply True and we can deduce the precondition for |tail xs| to be $|xs|\in\{(:)\}$. Substituting these into our precondition we find
\[ \textrm{True} \land (|null xs| \in \{|True|\} \lor |xs|\in\{(:)\}), \]
which simplifies to
\[ |null xs| \in \{|True|\} \lor |xs|\in\{(:)\}. \]
The subexpression  $|null xs| \in \{|True|\}$ forms a postcondition on |null xs|. From this postcondition Catch computes the required precondition that satisfies it, which in this particular case should turn out to be $|xs|\in\{[]\}$. Substituting, we find
\[ |xs|\in\{[]\} \lor |xs|\in\{(:)\}, \]
which, using our knowledge of the list data type, turns out the be a tautology, i.e. the (trivial) precondition for |safeTail| is True.

\subsection{Constraint Systems}
The example in the previous section used a simple constraint language, specifying to which set of head constructors an expression should evaluate. Mitchell developed two more expressive constraint systems: regular expression constraints and multipattern constraints.

\paragraph{Regular expression constraints} Constraints are formed by a regular expression over an alphabet of a number of ad-hoc \emph{selectors} (e.g. |hd| and |tl| for lists.) A precondition for |map head xs| using regular expression constraints reads $|xs| \in (|tl|^{*} \cdot |hd| \leadsto \{(:)\})$. It should be interpreted as ``after applying zero or more |tl|s to |xs| and then applying a |hd| we should find a |(:)| constructor.''

According to Mitchell regular expression constraints tend to scale badly as the program increase in size, although he could not identify more specific condition under which this problem manifests.\footnote{Personal communication}

\paragraph{Multipattern constraints} Multipatterns are of the form $\alpha \star \rho$, where $\alpha$ gives the set of constructors that are valid at the head of the value, while $\rho$ gives the set of constructors that can appear at the recursive positions (if any) of the constructor at the head of the value. The elements of these sets can again be multipatterns. To specify that |xs| in |map head xs| should be a list of non-empty lists, we use the multipattern:
\begin{align*}
\{[]&, (:) (\{(:)\ |Any| \} \star \{ [], (:)\ |Any|\})\} \\
&\star \\
\{[]&, (:) (\{(:)\ |Any| \} \star \{ [], (:)\ |Any|\})\}
\end{align*}
|Any| is a wildcard that matches any constructor.

Programmers define pre- and postconditions in the form of a contract (a refinement type). An appropriate contract for the |head| function defined above might be:
\begin{verbatim}
{-# CONTRACT head :: {s | not (null s)) -> {x | True} #-}
\end{verbatim}
This contract states that if head is given a list |s| for which the predicate |not (null s)| holds, i.e. it is a non-empty list, the function will not crash\footnote{The system only guarantees partial correctness, so the term might still diverge.}, indicated by the trivial postcondition |True| on the return value.

\begin{figure}
\centering
\begin{tabular}{lclr}
t &$::=$& $\{ x || p \}$     & Predicate          \\
  & $||$ & $x : t_1 \to t_2$ & Dependent function  \\
  & $||$ & $(t_1, t_2)$      & Constructor         \\
  & $||$ & $Any$             & Polymorphic ``Any''
\end{tabular}
\caption{Syntax of Xu's contracts}
\label{xu-contract-syntax}
\end{figure}

Contracts can be constructed from predicates, a dependent function space constructor, arbitrary constructors and a polymorphic |Any| contract that is always satisfied, including by functions which crash (Fig. \ref{xu-contract-syntax}). Any |Bool|-valued Haskell expression can be used as a predicate.

Dependent functions are helpful when declaring a contract, e.g. for the |reverse| function:
\begin{verbatim}
{-# CONTRACT reverse
      :: {xs | True} -> {rs | length xs == length rs} #-}
\end{verbatim}
The predicate on the return value depends on the input. Constructors and the |Any| contract are useful when declaring a contract for |fst|:
\begin{verbatim}
{-# CONTRACT fst :: (Ok, Any) -> Ok #-}
\end{verbatim}
Here |Ok| is used as a shorthand for |{x || True}|. As |fst| discards the value on the right in the tuple we should not care if it is an expression that crashes, so we can not use |Ok|.

\subsection{Contract Checking}
Haskell's syntax is extended with two \emph{exception values}---|BAD| and |UNR|---which are only used internally by the checker. |BAD| signifies an expression which crashes and |UNR| an expression which is either unreachable or diverges.

In the program that is going to be verified all crashing expressions (including |error| and |undefined|), as well as missing patterns in case-expressions are replaced by |BAD| exceptions.

The verification of the contracts is based on a translation of the contracts into Findler--Felleisen wrappers \cite{Findler:2002:CHF:581478.581484}, which will cause a function to crash (using |BAD|) if it disobeys its contract or diverge if it called in a way that is not permitted. While technically interesting it is not directly relevant to the detection of pattern match failures so we shall not discuss it in more depth.

Arithmetical expressions in case-expressions (e.g. |case x * x >= 0 of { True -> ...; False -> BAD }|) cannot be handled directly by the symbolic evaluator. These are collected in a table and send to an external constraint or SMT solver. If the solver determines these expressions are inconsistent the code is unreachable and can be pruned.

After the symbolic evaluation has terminated the checker will tell the programmer if the program is ``definitely satisfies the contract'' (if no |BAD| exceptions remain anywhere in the program), ``definitely does not satisfy the contract'' or ``don't know.''

Xi's initial example---recast in a more Haskell-like syntax---declares a |List| data type dependent on an integer corresponding to the length of the list and a function to concatenate two such lists:
\begin{verbatim}
type Nat = {a :: Int | a >= 0}

data List<Int> a = Nil<0>
                 | {n :: Nat} Cons<n+1> a (List<n> a)

(++) :: {m :: Nat} {n :: Nat} List<m> a -> List<n> a -> List<m+n> a
(++) Nil         ys = ys
(++) (Cons x xs) ys = Cons x (append xs ys)
\end{verbatim}

For many functions producing a list, the exact length might not be derived by such a trivial calculation (|m+n|) on the lengths of the input lists. A prominent example is the |filter| function:
\begin{verbatim}
filter :: (a -> Bool) -> {m :: Nat} List<m> a
                      -> [n :: Nat | n <= m] List<n> a
filter p Nil         = Nil
filter p (Cons x xs) | p x       = Cons x (filter p xs)
                     | otherwise =         filter p xs
\end{verbatim}
here we only know that the resulting list is equal or shorter in length than the input list (|n <= m|). This is expressed as the dependent sum |[n :: Nat || n <= m]|. In a full-fledged dependently-typed language we would also return a proof object stating the predicate |p| holds for all the elements in the output list, but this is beyond the expressive power of DML.

What we gain is a relief from the need to provide proofs for trivial arithmetical (in)equalities. Imagine a |zip| function which requires the two list to be zipped together to be of equal length. When calling |zip (xs ++ ys) (ys ++ xs)| this precondition seems to be intuitively satisfied. From the point-of-view of the compiler the first list has length $m+n$ and the second lists $n+m$, however. In most dependently-typed languages we will now have to invoke a lemma or tactic proving the commutativity of addition. DML can simply send the constraint $m + n = n + m$ to an ILP solver and conclude the constraint is satisfiable.

\subsection{Discussion}
Compared to Xu's static contract checking, Xi's DML constrains the constraint system to a decidable theory. The constraints and indices---such as |m+n| and |n <= m|---may superficially look like ordinary Haskell expression, but in fact belong to a much smaller index language. While type checking is decidable, type inference is not and as a result DML still requires type annotations.

To be applicable to a pattern match analysis we must try to infer a type like:
\begin{verbatim}
head :: {n :: Int || n >= 1} [a]<n> -> a
\end{verbatim}
for the |head| function. We also have to implicitly extend the list data type with an index representing its length.

This seems significantly more challenging than inferring the Catch-like type
\begin{alltt}
head :: \{xs :: [a] || xs \(\in\) (:) \} -> a
\end{alltt}
The correspondence between the type and the pattern matching happening in the definition of |head| is much more direct.

\section{Refinement Types}\label{secrt}

\begin{figure}[t]
\centering
\begin{tikzpicture}
    \tikzstyle{all nodes}=[inner sep=4pt]
    \draw node(top)at(1,3){|List a|}
          node(sorn)at(1,2){{|Singleton a|} $\lor$ {|Nil a|}}
          node(s)at(0,1){|Singleton a|} node(n)at(2,1){|Nil a|}
          node(bot)at(1,0){$\bot$};
    \draw[-](top)--(sorn);
    \draw[-](sorn)--(s);
    \draw[-](sorn)--(n);
    \draw[-](s)--(bot);
    \draw[-](n)--(bot);
\end{tikzpicture}
\caption{A lattice of {|List|} and its {|rectype|}s.}
\label{listlattice}
\end{figure}

Refinement types are a form of subtyping that allow us to state more accurately what values a variable of a particular type can contain. For example, a refinement type {\tt \{ a :: Int || a < 0 \}} states that |a| is a variable of type |Int| that can only contain negative numbers.

@@

Their key contribution is allowing the programmer to define |rectype|s. Besides defining a recursive type for lists:

\begin{verbatim}
data List a = Nil || Cons a (List a)
\end{verbatim}

the programmer can also define a |rectype| describing singleton lists:

@@

Together with an automatically derived |rectype| for the non-recurisive constructor |Nil| (and a union type contructor) the compiler can construct---using known algorithms on regular tree grammars \cite{DBLP:books/others/tree1984}---a finite lattice of refinement types for |List a| given in Figure \ref{listlattice}.

\subsection{Union and Intersection Types}\label{secit}

@@

A union type such as {\tt Int $\lor$ Bool}, {\tt List a $\lor$ Singleton a} or {\tt Int -> Int $\lor$ Bool -> Bool} state that a value can be either of the type on the left or of the type on the right, but we do not know which. In the case of {\tt List a $\lor$ Singleton a} we are able to simplify this type to {\tt List a}, as {\tt Singleton a} is a subtype of {\tt List a}.

@@

\begin{verbatim}
Cons  ::  a  -> Nil        a  -> Singleton  a
      &&  a  -> Singleton  a  -> List       a
      &&  a  -> List       a  -> List       a
\end{verbatim}

This type can be derived automatically from the |rectype| declarations. Types of functions can also be inferred automatically, although there may be a loss of precision when higher-order functions or polymorphism are involved.

@@

where |case_List| has the refinement type

@@

\section{Compiling Pattern Matching}
Case-statements in toy languages often have a very simple \emph{decision tree} semantics. Case-statements in Haskell have a more complex \emph{backtracking automaton} semantics. There is a body of work on compiling the latter into the former. Maranget \cite{DBLP:journals/jfp/Maranget07} gives an algorithm for determining whether a case-statement is exhaustive and whether all patterns are useful. As the analysis does not consider any dataflow information it is much too imprecise for our purpose. It does turn out that the analysis closely follows, but can at places be simplified with respect to, the manner in which pattern matching is compiled. This indicates we might also be able to follow such an approach when analyzing case-statements in our pattern-match analysis.
